---
title: "NNDSS Data Analysis"
output: html_notebook
---

# NNDSS Data Analysis

Exploratory Analysis of the National Notifiable Diseases Surveillance System (NNDSS) public dataset – [influenza (laboratory confirmed)](https://www.health.gov.au/resources/publications/national-notifiable-diseases-surveillance-system-nndss-public-dataset-influenza-laboratory-confirmed?language=en)

**Publication date:** September 2022



```{r}

library(tidyverse)
library(lubridate)
library(tsibble)
library(fable)
library(feasts)

lubridate.week.start = 1

source("scripts/ingest.R")
data <- ingest()

head(data)

```
**6 Variables**

- Week `date, ending on the Friday`
- State `NSW, NT, Qld, SA, Tas, Vic, WA`
- Age `bucketed in 5 year groups`
- Sex ` Male, Female, Unknown`
- Indigenous `Not available, Indigenous, Non-Indigenous, Unknown`
- Type `A(H1N1)pdm09, A(unsubtyped), A(H3N2), B, A and B, Untyped, C, A(H1N1)`

*Note: `A(H3N2)` are largely derived from PCR results reported as A(H3). Based on nationally available subtyping and characterisation data, these cases are assumed to be A(H3N2). `A and B` denotes a co-infection of both influenza A and influenza B viruses.*

## Caveats from source:

- These notification data represent only a proportion of the total cases occurring in the community, that is, only those cases for which health care was sought, a test conducted and a diagnosis made.
- In interpreting these data it is important to note that changes in notifications over time may not solely reflect changes in disease prevalence or incidence. Changes in testing policies; screening programs including the preferential testing of high-risk populations; the use of less invasive and more sensitive diagnostic tests; and periodic awareness campaigns, may influence the number of notifications that occur annually. 
- No ACT data
- Social distancing measures due to COVID causes discontinuities

[See full disclosure](https://www.health.gov.au/sites/default/files/documents/2022/09/national-notifiable-diseases-surveillance-system-nndss-public-dataset-influenza-laboratory-confirmed-caveats.pdf)


## Basic EDA

```{r}
data %>% count(State)
data %>% count(Age)
data %>% count(Sex)
data %>% count(Type)

```


*Thoughts*

- May want to normalise each state by the population at the time if doing statewise comparisons
- May want to cut off after 2020, due to covid-era capture issues
- Clear seasonality (ETS?)

```{r}
# accumulate for a tsibble

data %>%
  filter(Week < as.Date("2020-01-01 UTC")) %>%
  select(Week:Sex) %>% 
  group_by(Week,State,Age,Sex) %>% 
  summarise(Cases = n()) %>% ungroup() %>% 
  as_tsibble(index = Week, key = c(State, Age, Sex), regular = FALSE) -> datats

datats %>% summarise(Cases = sum(Cases)) %>% autoplot()

```
Decompose by year to show seasonality:

```{r}
datats %>%
  summarise(Cases = sum(Cases)) %>% 
  gg_season()
```




